SLC4A1 mutations that cause distal renal tubular acidosis alter cytoplasmic pH and cellular autophagy

Version 1

Abstract

Abstract

Distal renal tubular acidosis (dRTA) is a disorder characterized by the inability of the collecting duct system to secrete acids during metabolic acidosis. The pathophysiology of dominant or recessive SLC4A1 variant related dRTA has been linked with the mis trafficking defect of mutant kAE1 protein. However, in vivo studies in kAE1 R607H dRTA mice and humans have revealed a complex pathophysiology implicating a loss of kAE1-expressing intercalated cells and intracellular relocation of the H+-ATPase in the remaining type-A intercalated cells. These cells also displayed accumulation of ubiquitin and p62 autophagy markers. The highly active transport properties of collecting duct cells require the maintenance of cellular energy and homeostasis, a process dependent on intracellular pH. Therefore, we hypothesized that the expression of dRTA variants affect intracellular pH and autophagy pathways. In this study, we report the characterization of newly identified dRTA variants and provide evidence of abnormal autophagy and degradative pathways in mouse inner medullary collecting duct cells and kidneys from mice expressing kAE1 R607H dRTA mutant protein. We show that reduced transport activity of the kAE1 variants correlated with increased cytosolic pH, reduced ATP synthesis, attenuated downstream autophagic pathways pertaining to the fusion of autophagosomes and lysosomes and/or lysosomal degradative activity. Our study elucidated a close relationship between the expression of defective kAE1 proteins, reduced mitochondrial activity and decreased autophagy and protein degradative flux.

Introduction

Distal renal tubular acidosis (dRTA) is a disorder characterized by the inability of the collecting duct system to secrete acids during metabolic acidosis1. In addition to hyperchloremic metabolic acidosis, patients with this disease can present with hypokalemia, kidney stones, urinary sodium waste and difficulty to thrive. Expression of pathogenic variants in the ATP6V0A4, ATP6V1B1, FOXI1, WDR72 and SLC4A1 genes are the usual genetic aetiologies2–4. The SLC4A1 gene encodes the anion exchanger 1 (AE1) protein, which is an electroneutral chloride/bicarbonate exchanger5. It exists in two forms: a 911 amino acid erythroid isoform known to interact with erythroid cytoskeletal proteins and participate in red cell respiration and integrity, and a 65 amino acid (NH2-terminal) truncated isoform primarily found in the basolateral membrane of renal type A intercalated cells (A-IC)6,7 and podocytes8. This isoform participates in bicarbonate reabsorption and through its physical and functional interaction with the cytosolic carbonic anhydrase II and apical H+-ATPase, it supports apical proton export and urine acidification9.

Pathogenic variants in the SLC4A1 gene can result in either red cell defects (such as Southeast Asian ovalocytosis10,11 and hereditary spherocytosis12), renal cell defects (dRTA10) or both in patients with homozygous (Band 3 Coimbra and Band 3 Courcouronnes13,14) or compound heterozygous variants15,16. Renal SLC4A1 disease-causing variants have only been found in the transmembrane domain - where it could impact protein structure and its transport function- or in the short carboxyl domain – where it possibly affects protein-protein interactions. The pathophysiology of dominant or recessive SLC4A1 related dRTA (hereafter named dRTA) has originally been linked with the mis-trafficking defect of mutant kAE1 protein10,17. However, recent in vivo studies in mice and humans have revealed a complex pathophysiology where a loss of kAE1-expressing intercalated cells, the intracellular relocation of the H+-ATPase and accumulation of autophagy marker p62 and ubiquitin-positive material in the remaining type-A intercalated cells was suggested to be the principal cause of the disease18.

The highly active transport properties of collecting duct cells require the maintenance of cellular energy and homeostasis. The autophagy-mediated turnover of damaged organelles is necessary for protecting collecting duct cells as in most renal cells19. The chloride/bicarbonate exchange function of kAE1 in A-ICs confers a pivotal role in pH homeostasis and thus is a major contributor to cellular homeostasis. kAE1 protein interacts with several proteins such as integrin-like kinase (ILK), adaptor-related protein complex 1, 3 & 4 (AP-1, AP-3 & AP-4 mu1A), transmembrane protein 139 (TMEM139), kinesin family member 3B (KIF3B), and clathrin among others7,20–22 that support protein stability and trafficking. It also interacts with homeostatic proteins including the glycolytic enzyme glyceraldehyde-3-phosphate dehydrogenase (GAPDH)23 and the antioxidant enzyme peroxiredoxin 6 (PRDX 6)24, which play major roles in cellular energy metabolism and oxidative stress response, respectively.

In this study, we report the characterization of newly identified dRTA variations and provide evidence of abnormal autophagy and degradative pathways in cells and kidneys from mice expressing dRTA mutant kAE1 proteins.

Concise Methods

Ethics Approval

This study was conducted in accordance with all national and institutional animal care guidelines and approved by the University of Alberta’s Animal Care and Use Committee (AUP #1277).

Newly identified SLC4A1 variations from dRTA patients

The patient carrying the R295H mutation was a boy carrying the variation in the homozygous state, whose genetic diagnosis was made at the age of 5, following growth retardation of -2 SD for both weight and height, with bicarbonate at 18 mmol/L, potassium at 2.8 mmol/L, chloridemia at 94 mmol/L, calcemia at 2.55 mmol/L, and a urinary pH of 7.5. He also had a history of a pyeloureteral junction syndrome that was surgically managed at the age of 3. The R295H dRTA variation is a nonsense homozygous substitution characterized by a replacement of guanine (G) on position 884 by adenine (A) in the coding sequence. It has an allelic frequency of 0.14 % in the European population.

For the Y413H variant, the patient was a female diagnosed at 1 month old, with an urinary pH of 7.5, evidence of nephrocalcinosis and failure to thrive. The S525F variant has been previously reported25, but in brief, the patient was a 13 year old female with plasma pH of 7.25, plasma bicarbonate at 15.3 mmol/L who also presented with polyurethral junction syndrome, nephrocalcinosis and nephrolithiasis since childhood. The Y413H and S525F dRTA variations are nonsense heterozygous substitutions characterized by a replacement of thymine (T) at position 1237 by cytosine (C) and a replacement of cytosine (C) at position 1574 by thymine (T), respectively. The R589H dRTA variation has been previously described18. No follow up data were available for all patients.

Mice

Transgenic mice carrying the R607H knockin (KI) mutation (murine equivalent to human R589H mutation) were previously described18. Homozygous mice used throughout the study display incomplete dRTA with alkaline urine without metabolic acidosis at baseline as previously reported18. Homozygous mice or wild-type (WT) littermates were fed a standard rodent chow (Picolab® Rodent Diet 20 # 5053, LabDiet, ST. Louis, MO, USA) or for Figure 5 (H-K), a salt-depleted diet with acid challenge as previously reported26 with adequate and constant water supply, and maintained on a 12-hour light and dark cycle throughout their lifespan.

Statistical analysis

All the experiments were independently repeated a minimum of three times. Experimental results were analyzed using the GraphPad Prism software and are summarized as mean ± SEM. All statistical comparisons were made using unpaired Student’s t test or one/two-way ANOVA followed by a post-hoc test as indicated in figure legends. A P-value less than 0.05 was considered statistically significant. All datasets were assessed for normality, and outliers identified by Prism were excluded.

Detailed Material and Methods are listed in the Supplementary Material

Results

The dRTA kAE1 variants traffic to the basolateral membrane but have reduced transport activity in mIMCD3 cells

We first characterized three newly identified dRTA mutations and compared them with kAE1 WT or previously characterized kAE1 R589H mutant. Figure 1A depicts the alpha fold predicted structure of kAE1 showing amino acids mutated in each kAE1 mutant. kAE1 R295H is a recessively inherited substitution in the N-terminal cytosolic domain of the protein. kAE1 Y413H is a dominantly inherited substitution in transmembrane domain (TM) 1 of the core domain. In the gate domain, S525F and R589H substitutions occur in TM 5 and TM 6, respectively. We generated the newly identified dRTA variant cDNAs and expressed them or kAE1 WT in mIMCD3 cells to assess protein abundance (Figure 1B), localization (Figure 1C, D), transport activity (Figure 1E) and lifespan (Figure 1F). As seen on Figure 1B, both kAE1 R295H and S525F variants display 2 typical bands similar to kAE1 WT. The top band encompasses kAE1 proteins carrying a complex oligosaccharide that have reached the Golgi and beyond, while the bottom band corresponds to high mannose-carrying kAE1 proteins located in the endoplasmic reticulum. However, kAE1 Y413H mutant bands intensity was overall weaker than WT and displayed a predominant single band aligned with high mannose-carrying kAE1 proteins. These results indicate that the 3 newly described dRTA mutants are expressed in mIMCD3 cells. We next localized these mutants by immunofluorescence in mIMCD3 cells. While kAE1 WT is targeted to the plasma membrane, the kAE1 R295H, Y413H and S525F variants partially colocalize with endoplasmic reticulum marker calnexin (Figure 1C). Yet, both kAE1 WT and mutants appropriately co-localized with basolateral membrane marker beta-catenin in polarized mIMCD3 cells (Figure 1D). However, when the transport function was assayed on confluent cells, both kAE1 Y413H and S525F had significantly lower transport activity than WT (Figure 1E) and were not significantly different from un-induced WT-Dox cells (not shown on the graph). Finally, we observed that degradation of the kAE1 Y413H mutant begins 6 hours post-synthesis while kAE1 WT abundance remained stable for 24 hours (Figure 1F). Given this premature degradation, we did not perform further assays on cells expressing the Y413H protein. Overall, these results indicate that except for kAE1 Y413H mutant, the other mutants are expressed and traffic to the basolateral membrane of polarized mIMCD3 cells, similar to the previously published kAE1 R589H mutant.

kAE1 R295H, Y413H, S525F and R589H dRTA mutants are either dysfunctional or prematurely degraded.

(A) Alpha Fold predicted structure of the kidney isoform of Band 3 anion exchanger 1 (kAE1) with core and gate domains highlighted. The dRTA kAE1 mutation sites are coloured in blue with line extensions detailing specific amino acids mutated. (B) Immunoblot showing kAE1 expression and corresponding actin band of mIMCD kAE1 WT, R295H, Y413H, S525F and R589H cells treated with and without doxycycline for 24 hrs. Mouse anti-HA antibody was used to detect kAE1-HA, top and bottom bands correspond to kAE1 carrying complex and high mannose oligosaccharides, respectively. Immunostaining of kAE1 WT or mutant (red) and calnexin (green) in non-polarized mIMCD3 cells (C), and kAE1 WT or mutant (red) and b-catenin (green) in polarized (D) mIMCD cells. Scale bar = 10 μm. Red = kAE1, Green = ß-Catenin. (E) Rate of intracellular alkalinisation in WT or mutant mIMCD3 cells normalized to WT+ Dox. **** indicates P < 0.0001 using one-way ANOVA followed by a Dunnett’s post-hoc test. Error bars correspond to mean ± SEM, n= minimum 4. (F) Immunoblot of cycloheximide (CHX) chase assay with corresponding actin in kAE1 mIMCD WT and Y413H cells showing the degradation of kAE1 Y413H after 3hrs CHX incubation.

Expression of the kAE1 R295H, S525F and R589H dRTA mutants alter autophagy processes in mIMCD3 cells and whole kidney lysates

In mice expressing kAE1 R607H, a striking reduction in type-A intercalated cells was noted and in the remaining cells, autophagy marker p62 and ubiquitin accumulated in these abnormally enlarged cells18. We therefore investigated the autophagy machinery in dRTA mutant mIMCD3 cells and in R607H knockin (KI) mice. We first examined the ratio of autophagosome marker LC3BII protein relative to the total intensities of LC3BI and LC3BII as well as p62 levels. These experiments were performed at steady state, upon autophagy induction by starvation, or after autophagy inhibition by Bafilomycin (Baf) A1 (Figure 2A-C). Figure 2A – I show a consistent increase in the ratio of LC3B II to total LC3B (I + II) in the mutant cells at steady state (panel D), with starvation (panel F) and with Baf A1 (panel H) except for the kAE1 R295H mutant which was not significantly different from WT at steady state only. This suggests an altered autophagy process in the mutant cells, in agreement with preliminary findings from Mumtaz and colleagues18. There was no significant difference in p62 abundance in mutant cells compared to WT at steady state and with starvation (Figure 2E-G). However, with Baf A1, R589H mutant cells had significantly lower p62 abundance compared to WT (Figure 2I). To confirm these findings, we next assessed the abundance of these markers in whole kidney lysates from the kAE1 R607H KI mice. We observed that the total LC3B abundance was significantly higher in homozygous KI mice (KI/KI) compared to WT, with no difference in p62 (Figure 2J-L). Overall, these results suggest an abnormal autophagy in mutant mIMCD3 cells and in kidneys of R607H KI mice.

Autophagy is upregulated in dRTA kAE1 mutants in vitro and in vivo.

(A-C) Immunoblots of LC3 B and p62 with corresponding actin abundance in kAE1 WT, R295H, S525F and R589H mIMCD3 cells at steady state, under starvation (Starv) or 400 nM Bafilomycin A1 (Baf) treatment. (D-I) Quantification of immunoblots shown in A-C showing the ratio of LC3B II to total LC3B and p62. Error bars correspond to mean ± SEM, n= minimum 3. * P<0.05, ** P<0.01, *** P<0.005, ****P < 0.001 using one-way ANOVA followed by a Tukey’s post-hoc test. Immunoblots (J) and quantification (K, L) of LC3B and p62 abundance in kAE1 R607H KI mouse whole kidney lysates Error bars correspond to mean ± SEM, n= minimum 5. ***P < 0.005 using one-way ANOVA followed by a Tukey’s post-hoc test.

dRTA kAE1 mutant expressing cells have more alkaline intracellular pH than WT and altered autophagy flux

Given these preliminary findings of abnormal autophagy in the dRTA mutant cells, we next assessed the efficiency of the autophagy machinery in WT or dRTA mutant-expressing cells using the eGFP-RFP-LC3 construct (Figure 3A). The green fluorescent protein fused to LC3 is quenched in the acidic environment of the autolysosome, while both eGFP and RFP fluoresce in vesicles in the neutral lumen of the autophagosome27,28. Focusing on cells expressing either kAE1 S525F or R589H, we quantified the number of eGFP+ (green), RFP+ (red) and double eGFP+/RFP+ (yellow) vesicles in the kAE1 mutant or WT transfected live cells (Figure 3B-D). kAE1 S525F mutant cells have significantly more autophagosomes (yellow) than WT counterparts (Figure 3B), and both kAE1 S525F and R589H have significantly more autolysosomes (red) than WT (Figure 3C). Figure 3D shows that both mutants, particularly the kAE1 S525F mutant cells had significantly more autophagosomes than autolysosomes. This finding suggests an upregulation of autophagy and inhibition in the downstream steps of autophagy that involves the fusion of the autophagosome with the lysosome29. As autolysosomes require luminal v-H+-ATPase -dependent acidification to efficiently clear cell debris 30 and as kAE1 is a base transporter whose function is linked to the v-H+-ATPase, we next wondered whether the steady-state cytosolic pH of mIMCD3 cells expressing kAE1 WT or mutant was different. Using BCECF-AM, we observed that the steady state intracellular pH (pHi) of kAE1 mutant cells was more alkaline than WT cells (except for kAE1 R295H cell pHi which has a similar trend but is not significantly different from WT) (Figure 3E), in agreement with transport activity (Figure 1F). Given the lack of difference in phenotype between the kAE1 R295H and kAE1 WT mIMCD3 cells, we focused the subsequent experiments on kAE1 S525F and R589H mutant cells. As a higher intracellular pH could contribute to alterations in the autophagy machinery31, we chemically acidified the cytosolic pH of cells expressing kAE1 mutant or WT and examined autophagy and lysosomal markers (Figure 3 F-H). We first determined that incubation of mIMCD3 cells in 0.033 μM nigericin for 2 hours acidified cytosolic pH to 6.9 without causing cell death (Supplementary Figure 1). We observed that chemically reducing pHi to 6.9 in mutant expressing cells reduced the ratio of LC3B II to total LC3B (Figure 3I) and the abundance of lysosomal-associated membrane protein 1 (LAMP1) in R589H cells (Figure 3 J) to levels similar to WT cells at steady state. These findings suggest that abnormal autophagy in the mutant cells may be caused by their alkaline pHi, resulting from a reduced anion exchange activity of the mutant kAE1 protein.

dRTA kAE1 mutants have more alkaline steady-state intracellular pH and altered autophagy flux.

(A) Immunofluorescence staining of kAE1 in eGFP-RFP-LC3 transfected mIMCD3 cells expressing kAE1. GFP = green, RFP = red, kAE1= cyan. Scale bar = 8 μm. Graphical representation of number of yellow (autophagosomes) (B) and red (autolysosomes) (C) puncta per cell expressing kAE1. Error bars correspond to mean ± SEM, n= minimum 32. ** P<0.01, *** P<0.005, ****P < 0.001 using one-way ANOVA followed by a Tukey’s post-hoc test. (D) Grouped graph of the number of yellow (autophagosomes) and red (autolysosomes) puncta per cell expressing kAE1 respectively. Note that the statistical analysis displayed only compared yellow and red groups for simplification. Error bars correspond to mean ± SEM, n= minimum 32. ** P<0.01, ****P < 0.001 using two way ANOVA followed by a Sidak’s post-hoc test. (E) Graphical representation of intracellular pH measurement of mIMCD kAE1 WT, R295H, Y413H, S525F and R589H cells. Error bars correspond to mean ± SEM, n = minimum 32. *P<0.05, **P<0.01 using one-way ANOVA followed by a Dunnett’s post-hoc test. (F-H) Immunoblot of LC3B, LAMP1 and actin in kAE1 WT, S525F and R589H mIMCD3 cells at steady state and under chemically reduced intracellular pH conditions. Graphical representation of the ratio of LC3B II to total LC3B ratio (I) or LAMP 1 (J) at steady state versus at low pHi in mIMCD3 kAE1 WT, S525F and R589H. Black circles indicate steady state cells and triangles indicate low pHi cells. Error bars correspond to mean ± SEM, n= 3. ** indicates P<0.01 using two-way ANOVA followed by a Sidak’s post-hoc test.

mIMCD3 cells expressing dRTA kAE1 mutants and R607H KI kidney tissues have abnormal lysosome number and size

The accumulation of autophagosomes and autolysosomes as seen above may suggest one or a combination of the following: an inability of autophagosomes to fuse with lysosomes and/ or a defect in lysosomal degradative activity in the mutant cells19,32. We first examined the lysosomal degradative activity by assessing lysosomal protease Cathepsin B activity using Magic Red staining (Figure 4A-B). In agreement with increased RFP+ vesicles (Figure 3D), the kAE1 S525F mutant cells had a significantly higher number of Magic Red positive vesicles than WT whereas the kAE1 R589H mutant cells had significantly larger Magic Red positive vesicles, suggesting an accumulation of undigested material33. To validate this finding, we performed immunostaining and quantified LAMP1 positive staining in ß1 ATPase positive cells (a marker of ICs) in WT and R607H KI mouse kidney sections. The KI mice showed significantly more and larger LAMP1 positive vesicles compared to WT mice in both cortex and medulla (Figure 4 C-F). We probed further into the lysosomal activity by quantifying lysosomal protease Cathepsin D (immature, intermediate and mature) protein abundance by immunoblot in isolated primary murine A-ICs. Although the abundance of immature and intermediate cathepsin D did not differ between genotypes, the KI mice showed a significantly decreased abundance of mature cathepsin D (Supplementary Figure 2). Thus, in line with in vitro findings, A-IC from homozygous R607H KI mice display relatively more and larger lysosomes with reduced active protease abundance than WT littermates, suggesting a lysosomal defect in the dRTA kAE1 mutant cells.

dRTA kAE1 mutants have bigger or more lysosomes than WT in vitro and in vivo.

(A) Immunofluorescence images of kAE1 WT, S525F and R589H mIMCD3 cells incubated with Magic Red substrate for 1hr at 37°C in the dark. Green = kAE1, magenta = active lysosomes, blue = nuclei. Scale bar =16 µm. (B) Graphical representation of number and size of active lysosomes per cell. Error bars correspond to mean ± SEM, n= minimum 30. **** P<0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. Immunofluorescence images of LAMP 1 and ß1 ATPase in kidney cortex (C) or medulla (E) from kAE1 WT and R607H KI mice. Blue = nuclei, magenta = LAMP 1 (lysosomes), Yellow = ß1 ATPase. Scale bar = 8um. Graphical representation of the number and volume of LAMP1 vesicles in ß1 ATPase positive cells in the kidney cortex (D) or medulla (F) of WT or R607H KI mice. Error bars correspond to mean ± SEM, n= 60. **** P<0.001 using Student’s t-test.

dRTA kAE1 mutant cells have lower ATP production rate and abnormal mitochondrial content.

(A - B) Oxygen consumption rate (OCR) and Extra Cellular Acidification Rate (ECAR) of empty vector transfected cells, kAE1 WT, S525F or R589H mIMCD3 cells analyzed in a Seahorse XFe96 Extracellular Flux Analyzer with the ATP Rate Assay Test Kit. The empty vector transfected cells provided a control for a potential effect of doxycycline on measurements. (C) Graphical representation of the combination of ATP production rates from mitochondrial respiration (mitoATP) and glycolysis (glycoATP) of kAE1 WT, S525F and R589H mIMCD3 cells measured in real-time following sequential injections of oligomycin and Rotenone +Antimycin A. Error bars correspond to mean ± SEM, n= minimum. * p < 0.05, ****p < 0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. Graphical representations of mitochondrial respiration (D) and glycolytic ATP production (E) in kAE1 WT, S525F and R589H mIMCD3 cells. Error bars correspond to mean ± SEM, n= minimum 8. **P<0.01, ****p < 0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. (F) Immunofluorescence staining of TOM 20 and kAE1 in kAE1 WT, S525F and R589H mIMCD3 cells. Blue = nuclei, Magenta = TOM 20, Green = kAE1. Scale bar = 8 µm. (G) Graphical representation of total TOM 20 fluorescence intensity per cell expressing kAE1. Error bars correspond to mean ± SEM, n= minimum 40. ***p < 0.001, ****p < 0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. Immunofluorescence images of TOM 20 and ß1 ATPase in kidney cortex (H) or medulla (I) of kAE1 R607H WT and KI mice exposed to a salt depleted diet with an acid challenge26. Blue = nuclei, magenta = TOM 20 (mitochondria), Yellow = ß1 ATPase. Scale bar = 8um. (I) Graphical representation of the total TOM 20 fluorescence intensity in ß1 ATPase positive cells in the cortex (J) or medulla (K) of the kidney. Error bars correspond to mean ± SEM, n= 90. * p < 0.05, ****p < 0.0001 using Student’s t-test.

dRTA kAE1 mutant cells have lower ATP production rate and abnormal mitochondrial content

Lysosomal degradation is highly dependent on a low luminal pH generated in part by the vacuolar-type H+-ATPase34 whose activity depends on ATP hydrolysis. We therefore analysed the ATP production rate in mIMCD3 cells, specifically glycolysis and oxidative phosphorylation. We measured the oxygen consumption rate (OCR) (Figure 5A) and extracellular acidification rate (ECAR) in empty vector transfected, kAE1 WT or mutant cells (Figure 5B). Both kAE1 S525F and R589H mutant cells had a lower ATP production rate compared to WT (Figure 5C). More specifically, the R589H mutant cells had a lower mitochondrial ATP production rate (Figure 5D) whereas the kAE1 S525F mutant cells exhibited a lower glycolytic ATP production rate (Figure 5E). With the mitochondria being the major ATP-producing organelles35, we assessed mitochondrial content by immunostaining of translocase of the outer membrane 20 (TOM20) both in vitro and in vivo. Both kAE1 S525F and R589H mutant cells have higher mitochondrial content compared to WT as determined by total overall intensity of TOM20 positive puncta (Figure 5F&G). In line with this result, although a decreased fluorescence intensity was observed in the cortex, there was a significantly higher TOM20 fluorescence intensity in medullary kidneys of homozygous R607H KI mutant mice compared to WT littermates (Figure 5H-K).

Discussion

In this study, we characterized three newly identified dRTA-causing kAE1 variations. Combining in vivo and in vitro studies, we demonstrated that reduced transport activity of the kAE1 mutants correlated with increased cytosolic pH, reduced ATP synthesis, attenuated downstream autophagic pathways and lysosomal dysfunction, pertaining to the fusion of autophagosomes and lysosomes and/ or lysosomal degradative activity.

In line with previous observations18, the kAE1 R295H, Y413H and S525F mutants were intracellularly retained in non-polarised mIMCD3 cells but properly localized to the basolateral membrane after polarization. When expressed in Madin-Darby canine kidney (MDCK) cells, other dRTA-causing kAE1 mutants such as dRTA R602H, G701D, V488M, deltaV850 variants also exhibited a plasma membrane trafficking defect10,36,37. In contrast, the kAE1 R589H mutant was correctly targeted to the plasma membrane18. Functionally, cells expressing the Y413H and S525F mutants exhibit about 40 % reduction in chloride/bicarbonate exchange activity compared to kAE1 WT, similar to the previously characterized recessive G701D mutant but unlike the R295H mutant38. Therefore, the mechanism causing dRTA remains unclear in the case of the newly identified R295H variant. These findings add to the growing list of SLC4A1 gene variations causing dRTA.

We next investigated the roots of the altered autophagy briefly reported in R607H KI mice18 using mIMCD3 cells and whole kidney lysates. In kidneys of the KI mice, a decrease in A-IC, accumulation of p62 and ubiquitin and enlarged remaining A-ICs, prompted us to investigate autophagy pathways in dRTA mutant cells and in homozygous R607H KI mice18,38. During the autophagy process, LC3B I (a marker for autophagosomes) is converted to lipidated LC3B II39 and p62 aggregates to facilitate the degradation of ubiquitinated proteins within the autophagosome complex40. In mIMCD3 cells, LC3B lipidation was increased in the kAE1 R295H, S525F and R589H mutant cells compared to WT, an increase that persisted with both Baf A1 treatment and starvation. Although opposite effects were expected under inhibition or induction of autophagy, such similar effect has been previously described. In the proximal tubule of obese mice, LC3B accumulation indicating a stagnated autophagy flux, was observed with both chloroquine treatment and 24 hour starvation41. Similarly, in NRK-52E cells, a disruption of the autophagy machinery by LC3B II accumulation in high cadmium-stressed cells occurred under either Baf A1 or rapamycin (an autophagy inducer) treatment42. Although LC3B II is elevated during both increased autophagy flux and disrupted autophagy, we did not observe significant accumulation of p62 in mIMCD3 cells expressing dRTA mutants. p62 is specifically a marker of autophagy-mediated protein clearance43–45. Therefore, p62 accumulation in R607H KI mouse kidney sections strongly suggests a compromised autophagy-mediated clearance while the increased LC3B lipidation without significant changes in p62 in mIMCD3 cells points towards either an increased autophagic flux and/or a disrupted autophagy.

To obtain a clearer picture of the precise autophagic pathway altered in the dRTA mutants, we probed further into the different stages of autophagy and autophagy flux. We noted an accumulation of autophagosomes and autolysosomes in the S525F and R589H mutant cells. This was recapitulated in the R607H KI mice which showed significantly more and bigger LAMP 1 positive vesicles in both kidney cortex and medulla, suggesting a blockage in autophagy flux in both dRTA mutant cells and KI mice. Such blockage has been implicated in the pathophysiology of several diseases. In lysosomal storage disease, lysosome accumulation in proximal tubule cells is a key component in the pathways mediating epithelial dysfunction19. In this study, restoring autophagy flux attenuated disease progression. Another study in SK-N-SH, RT4-D6P2T and HeLa cells implicated autophagosome and lysosome accumulation in cellular toxicity associated with neurodegenerative diseases32. In agreement with altered lysosomal function, we also noted a greater abundance and size of active cathepsin B lysosomal protease vesicles in mIMCD3 cells. Increased cathepsin B activity affects lysosomal biogenesis, autophagy initiation and cellular homeostasis46. In the renal context, cathepsin B knockout mice demonstrated a higher resistance and quicker recovery from glomerular damage47, suggesting that cathepsin B accumulation may be detrimental to the cells. Increased cathepsin B abundance in the dRTA mutant cells also correlates with the accumulation of lysosomes. Overall, these findings suggest that the pathogenesis of dRTA in our models involves an inhibition of autophagy flux at the downstream steps involving autophagosome-lysosome fusion and lysosomal protein clearance 29,33,48,49.

The question remained as to how these dRTA variants altered autophagy. The SLC4A1-3 gene family that includes AE1 are regulators of intracellular pH in different cell types50–52. The reduced anion exchange activity in mIMCD3 cells expressing the R589H and S525F variants expectedly correlated with an increased pHi compared to WT counterparts. We wondered whether this was the mechanism explaining blockade in autophagy flux since pHi variations impact autophagy31,34,53–55. Previous studies reported more perinuclear localization of lysosomes and autophagosome-lysosome fusion in cells with an increased pHi53,54. Although not examined in our study, an increase in intracellular pH due to starvation decreased the levels of lysosomal kinesin superfamily member KIF2 and ADP-ribosylation factor-like 8B (ARL8), which are responsible for redistributing lysosomes to the cell periphery. This reduction subsequently inhibited mTORC1, resulting in increased autophagosome synthesis and autophagosome-lysosome fusion53.

While an alkaline cytosolic pH partially explains the impairment in autophagy flux, a close relationship also exists between intracellular pH and cellular energy dynamics and metabolic stress, all of which are key regulators of autophagy56. Therefore, it was plausible that kAE1 variant-induced autophagy dysregulation occurs through a signalling pathway akin to that of energy deprivation-induced autophagy. This hypothesis is supported by our findings of reduced ATP production rate in dRTA kAE1 S525F and R589H mutant cells compared to kAE1 WT cells. We also found that both kAE1 S525F and R589H mutant cells have higher mitochondrial content compared to kAE1 WT cells. The increased mitochondrial content coupled with low ATP point towards improper mitochondrial function in dRTA variant cells. Low ATP levels as seen in dRTA mutant cells may impair autophagy as was shown in human RPE cells57. ATP reduction in RPE cells led to complex mitochondrial changes such as structural disorganization, enzyme activity decline, and oxidative damage to mitochondrial components and DNA57. Similarly, in pancreatic islet cells, an alkaline pHi led to increased uptake of phosphate by mitochondria, accelerating the production of superoxide, promoting mitochondrial permeability transition, and inducing translational attenuation due to endoplasmic reticulum stress, ultimately impairing insulin secretion58. Overall, our data support that expression of the kAE1 variants increases pHi which alters mitochondrial function and leads to reduced cellular energy levels that eventually attenuates energy-dependent autophagic pathways including autophagosome-lysosome fusion and lysosomal protein clearance.

In light of these observations, we postulated that correcting the alkaline pHi of dRTA mutant-expressing mIMCD3 cells will alleviate this blockage in autophagy flux. We observed that a chemically engineered pHi of 6.959 reduced LC3B II accumulation and LAMP1 abundance in mIMCD3 mutant cells to expression levels similar to that of WT cells at baseline. This suggests that a chemically reduced pHi facilitated protein clearance in the two dRTA mutant cells31,53 as noted in other studies. In one such study, treatment of SH-SY5Y cells with FCCP and nigericin also acidified intracellular pH and triggered autophagy and mitophagy31. Similarly, acid loading in proximal tubular cells under chronic metabolic acidosis showed increased autophagic flux and mitophagy60. These findings are in line with our results and establish a link between altered autophagy flux and the alkaline pHi of dRTA variant cells.

In conclusion, our study established a strong relationship between the expression of defective kAE1 proteins, reduced mitochondrial activity, decreased autophagy and impaired protein degradative flux. Whether this abnormal degradative pathway explains the premature loss of A-IC will need to be elucidated in further studies.

eLife Assessment

This work characterizes the function and localization of SLC4A1 variants associated with distal renal tubular acidosis in human patients. Cell culture and limited animal studies provide partial but incomplete support to the authors' claim that the variants disrupt normal protein degradative flux by alkalinizing the intracellular pH. The study is valuable in providing preliminary evidence for future exploration of the link between intracellular pH regulation by SLC4A1 and kidney cell function in vivo.

Reviewer #1 (Public review):

Summary:

This study is an evaluation of patient variants in the kidney isoform of AE1 linked to distal renal tubular acidosis. Drawing on observations in the mouse kidney, this study extends findings to autophagy pathways in a kidney epithelial cell line.

Strengths:

Experimental data are convincing and nicely done.

Weaknesses:

Some data are lacking or not explained clearly. Mutations are not consistently evaluated throughout the study, which makes it difficult to draw meaningful conclusions.

Reviewer #2 (Public review):

Context and significance:

Distal renal tubular acidosis (dRTA) can be caused by mutations in a Cl-/HCO3- exchanger (kAE1) encoded by the SLC4A1 gene. The precise mechanisms underlying the pathogenesis of the disease due to these mutations are unclear, but it is thought that loss of the renal intercalated cells (ICs) that express kAE1 and/or aberrant autophagy pathway function in the remaining ICs may contribute to the disease. Understanding how mutations in SLC4A1 affect cell physiology and cells within the kidney, a major goal of this study, is an important first step to unraveling the pathophysiology of this complex heritable kidney disease.

Summary:

The authors identify a number of new mutations in the SLC4A1 gene in patients with diagnosed dRTA that they use for heterologous experiments in vitro. They also use a dRTA mouse model with a different SLC4A1 mutation for experiments in mouse kidneys. Contrary to previous work that speculated dRTA was caused mainly by trafficking defects of kAE1, the authors observe that their new mutants (with the exception of Y413H, which they only use in Figure 1) traffic and localize at least partly to the basolateral membrane of polarized heterologous mIMCD3 cells, an immortalized murine collecting duct cell line. They go on to show that the remaining mutants induce abnormalities in the expression of autophagy markers and increased numbers of autophagosomes, along with an alkalinized intracellular pH. They also reported that cells expressing the mutated kAE1 had increased mitochondrial content coupled with lower rates of ATP synthesis. The authors also observed a partial rescue of the effects of kAE1 variants through artificially acidifying the intracellular pH. Taken together, this suggests a mechanism for dRTA independent of impaired kAE1 trafficking and dependent on intracellular pH changes that future studies should explore.

Strengths:

The authors corroborate their findings in cell culture with a well-characterized dRTA KI mouse and provide convincing quantification of their images from the in vitro and mouse experiments.

Weaknesses:

The data largely support the claims as stated, with some minor suggestions for improving the clarity of the work. Some of the mutants induce different strengths of effects on autophagy and the various assays than others, and it is not clear why this is from the present manuscript, given that they propose pHi and the unifying mechanism.

Reviewer #3 (Public review):

Summary:

The authors have identified novel dRTA causing SLC4A1 mutations and studied the resulting kAE1 proteins to determine how they cause dRTA. Based on a previous study on mice expressing the dRTA kAE1 R607H variant, the authors hypothesize that kAE1 variants cause an increase in intracellular pH, which disrupts autophagic and degradative flux pathways. The authors clone these new kAE1 variants and study their transport function and subcellular localization in mIMCD cells. The authors show increased abundance of LC3B II in mIMCD cells expressing some of the kAE1 variants, as well as reduced autophagic flux using eGFP-RFP-LC3. These data, as well as the abundance of autophagosomes, serve as the key evidence that these kAE1 mutants disrupt autophagy. Furthermore, the authors demonstrate that decreasing the intracellular pH abrogates the expression of LC3B II in mIMCD cells expressing mutant SLC4A1. Lastly, the authors argue that mitochondrial function, and specifically ATP synthesis, is suppressed in mIMCD cells expressing dRTA variants and that mitochondria are less abundant in AICs from the kidney of R607H kAE1 mice. While the manuscript does reveal some interesting new results about novel dRTA causing kAE1 mutations, the quality of the data to support the hypothesis that these mutations cause a reduction in autophagic flux can be improved. In particular, the precise method of how the western blots and the immunofluorescence data were quantified, with included controls, would enhance the quality of the data and offer more supportive evidence of the authors' conclusions.

Strengths:

The authors cloned novel dRTA causing kAE1 mutants into expression vectors to study the subcellular localization and transport properties of the variants. The immunofluorescence images are generally of high quality, and the authors do well to include multiple samples for all of their western blots.

Weaknesses:

Inconsistent results are reported for some of the variants. For example, R295H causes intracellular alkalinization but also has no effect on intracellular pH when measured by BCECF. The authors also appear to have performed these in vitro studies on mIMCD cells that were not polarized, and therefore, the localization of kAE1 to the basolateral membrane seems unlikely, based upon images included in the manuscript. Additionally, there is no in vivo work to demonstrate that these kAE1 variants alter intracellular pH, including the R607H mouse, which is available to the authors. The western blots are of varying quality, and it is often unclear which of the bands are being quantified. For example, LAMP1 is reported at 100kDa, the authors show three bands, and it is unclear which one(s) are used to quantify protein abundance. Strikingly, the authors report a nonsensical value for their quantification of LCRB II in Figure 2, where the ratio of LCRB II to total LCRB (I + II) is greater than one. The control experiments with starvation and bafilomyocin are not supportive and significantly reduce enthusiasm for the authors' findings regarding autophagy. There are labeling errors between the manuscript and the figures, which suggest a lack of vigilance in the drafting process.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study is an evaluation of patient variants in the kidney isoform of AE1 linked to distal renal tubular acidosis. Drawing on observations in the mouse kidney, this study extends findings to autophagy pathways in a kidney epithelial cell line.

Strengths:

Experimental data are convincing and nicely done.

Thank you

Weaknesses:

Some data are lacking or not explained clearly. Mutations are not consistently evaluated throughout the study, which makes it difficult to draw meaningful conclusions.

We have revised our manuscript to clarify some earlier explanations and provided rationale for focusing on specific variants throughout the study.

Reviewer #2 (Public review):

Context and significance:

Distal renal tubular acidosis (dRTA) can be caused by mutations in a Cl-/HCO3- exchanger (kAE1) encoded by the SLC4A1 gene. The precise mechanisms underlying the pathogenesis of the disease due to these mutations are unclear, but it is thought that loss of the renal intercalated cells (ICs) that express kAE1 and/or aberrant autophagy pathway function in the remaining ICs may contribute to the disease. Understanding how mutations in SLC4A1 affect cell physiology and cells within the kidney, a major goal of this study, is an important first step to unraveling the pathophysiology of this complex heritable kidney disease.

Summary:

The authors identify a number of new mutations in the SLC4A1 gene in patients with diagnosed dRTA that they use for heterologous experiments in vitro. They also use a dRTA mouse model with a different SLC4A1 mutation for experiments in mouse kidneys. Contrary to previous work that speculated dRTA was caused mainly by trafficking defects of kAE1, the authors observe that their new mutants (with the exception of Y413H, which they only use in Figure 1) traffic and localize at least partly to the basolateral membrane of polarized heterologous mIMCD3 cells, an immortalized murine collecting duct cell line. They go on to show that the remaining mutants induce abnormalities in the expression of autophagy markers and increased numbers of autophagosomes, along with an alkalinized intracellular pH. They also reported that cells expressing the mutated kAE1 had increased mitochondrial content coupled with lower rates of ATP synthesis. The authors also observed a partial rescue of the effects of kAE1 variants through artificially acidifying the intracellular pH. Taken together, this suggests a mechanism for dRTA independent of impaired kAE1 trafficking and dependent on intracellular pH changes that future studies should explore.

Strengths:

The authors corroborate their findings in cell culture with a well-characterized dRTA KI mouse and provide convincing quantification of their images from the in vitro and mouse experiments

Thank you

Weaknesses:

The data largely support the claims as stated, with some minor suggestions for improving the clarity of the work. Some of the mutants induce different strengths of effects on autophagy and the various assays than others, and it is not clear why this is from the present manuscript, given that they propose pHi and the unifying mechanism

We have modified our manuscript to discuss the various strengths of the mutants and emphasize that alteration of cytosolic pH by kAE1 variants may not be the only mechanism leading to dRTA.

Reviewer #3 (Public review):

Summary:

The authors have identified novel dRTA causing SLC4A1 mutations and studied the resulting kAE1 proteins to determine how they cause dRTA. Based on a previous study on mice expressing the dRTA kAE1 R607H variant, the authors hypothesize that kAE1 variants cause an increase in intracellular pH, which disrupts autophagic and degradative flux pathways. The authors clone these new kAE1 variants and study their transport function and subcellular localization in mIMCD cells. The authors show increased abundance of LC3B II in mIMCD cells expressing some of the kAE1 variants, as well as reduced autophagic flux using eGFP-RFP-LC3. These data, as well as the abundance of autophagosomes, serve as the key evidence that these kAE1 mutants disrupt autophagy. Furthermore, the authors demonstrate that decreasing the intracellular pH abrogates the expression of LC3B II in mIMCD cells expressing mutant SLC4A1. Lastly, the authors argue that mitochondrial function, and specifically ATP synthesis, is suppressed in mIMCD cells expressing dRTA variants and that mitochondria are less abundant in AICs from the kidney of R607H kAE1 mice. While the manuscript does reveal some interesting new results about novel dRTA causing kAE1 mutations, the quality of the data to support the hypothesis that these mutations cause a reduction in autophagic flux can be improved. In particular, the precise method of how the western blots and the immunofluorescence data were quantified, with included controls, would enhance the quality of the data and offer more supportive evidence of the authors' conclusions.

Strengths:

The authors cloned novel dRTA causing kAE1 mutants into expression vectors to study the subcellular localization and transport properties of the variants. The immunofluorescence images are generally of high quality, and the authors do well to include multiple samples for all of their western blots.

Thank you

Weaknesses:

Inconsistent results are reported for some of the variants. For example, R295H causes intracellular alkalinization but also has no effect on intracellular pH when measured by BCECF. The authors also appear to have performed these in vitro studies on mIMCD cells that were not polarized, and therefore, the localization of kAE1 to the basolateral membrane seems unlikely, based upon images included in the manuscript. Additionally, there is no in vivo work to demonstrate that these kAE1 variants alter intracellular pH, including the R607H mouse, which is available to the authors. The western blots are of varying quality, and it is often unclear which of the bands are being quantified. For example, LAMP1 is reported at 100kDa, the authors show three bands, and it is unclear which one(s) are used to quantify protein abundance. Strikingly, the authors report a nonsensical value for their quantification of LCRB II in Figure 2, where the ratio of LCRB II to total LCRB (I + II) is greater than one. The control experiments with starvation and bafilomyocin are not supportive and significantly reduce enthusiasm for the authors' findings regarding autophagy. There are labeling errors between the manuscript and the figures, which suggest a lack of vigilance in the drafting process.

The R295H variant was identified in a dRTA patient and as such, it was important to report it. However, this is the first mutation located in the amino-terminus of the protein, which may be involved in protein-protein interactions, so other mechanisms may cause dRTA for this variant. We have therefore modified our manuscript to state that alteration of cytosolic pH may not be the only mechanism leading to dRTA. At this time, we are not able to measure cytosolic pH in vivo and hope to be able to do it in the future.

In our revised manuscript, we also show cell surface biotinylation results supporting that plasma membrane abundance of the kAE1 S525F and R589H variants is not significantly different than WT in non-polarized mIMCD3 cells (Figure 3 A&B), in line with the predominant basolateral localization of the variants in polarized cells (Figure 1C). Therefore, these two mutant proteins are not mis-trafficked in non-polarized cells. Finally, we have clarified which bands have been used for quantification and corrected quantifications (including ratio measurements).

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) R295H is recessively inherited, whereas Y413H is dominantly inherited: this is interesting and may be linked to their cellular expression and function. Is this information known for the other mutations examined in this study?

The S25F and R589H dRTA variants have both been reported to exhibit autosomal dominant inheritance. This information is now updated in lines 146 and 158-159.

(2) R589H expression levels are evaluated in the Western blot of Figure 1, but localization and activity are not examined in Figure 2. However, R589H is included in autophagy experiments shown in later figures. Similarly, mutant R607H is the subject of several experiments further into the manuscript, but no initial analysis is provided for this variant.

Protein abundance and localization of the R589H mutant in mIMCD3 cells have been shown in our previous publication in Supplementary Fig 5D and Supplementary Fig 2J [1]. This now indicated on lines 158-159. Our previous paper also presented a detailed study of the R607H dRTA mutant, the mouse model corresponding to the human R589H mutation. This is now indicated on lines 70, 118-119 and 180. The present study builds upon those published findings.

(3) This inconsistency is confusing, detracts from the usefulness of the study, and makes the comparative analysis of mutations incomplete. It is difficult to extrapolate from published studies in MDCK1 cells, which show different results on trafficking.

The mIMCD3 cell line, which more closely resembles the physiology of the mouse collecting duct than MDCK cells, was selected for this study and our previous one [1]. Accordingly, the results obtained are better aligned with in vivo evidence. In contrast, differences in mutant protein expression and localization observed in other cell lines, like the MDCK cells, are likely attributable to differences in their cellular origin.

(4) In Figure 2, could the authors explain why total LC3B is graphed for the data shown in mouse lysates, whereas the ratio of bands is analysed for cell lysates? Both sets of data show the two LC3B bands.

Total LC3B levels were significantly increased in the mutant compared to WT; however, no significant difference was observed in the lipidation ratio. For this reason, that graph is not shown in the main paper but has been included in the Supplementary Figure 1D.

(5) In Figure 3, representative fluorescence images should be shown for all cell lines.

We have now included representative immunofluorescence images for all cell lines in Figure 3C.

(6) pH effects: Suggest that steady state pHi (Figure 3E) and rate of alkalization (Figure 1F) would be more effective together in Figure 1. The authors should show data for the effect of nigericin on cytoplasmic pH in Figure 3. If the rate of alkalinization in the mutant cells is reduced, shouldn't the intracellular steady state pH be more acidic? A cartoon depicting the transporter activity in the cell and the expected changes in pHi would be helpful. Is there a way to activate/inhibit NHE1 and rescue the effect of the mutant kAE1? It is unclear if the link between the mutant kAE1 and mitochondrial ATP production is a consequence of the intracellular pH or an indirect effect.

We opted to keep the effect of nigericin on pHi in Supplementary Fig1A given that Figure 3 already contains 11 panels. Also, in intercalated cells, the kAE1 protein physiologically exports 1 molecule of bicarbonate in exchange of 1 chloride ion import hence a reduced transport activity would result in a more alkaline intracellular pH. To clarify this point, we have included a diagram in Figure 1E as suggested. However, to calculate the rate of intracellular alkalinisation, the transporter is functioning in the opposite direction, i.e. extruding chloride and importing bicarbonate (see methods protocol for transport assay). Therefore, in this assay (Figure 1G), a defective chloride/bicarbonate activity results in a reduced rate of intracellular alkalinisation rate. This is now explained on lines 169-172.

Disruption of NHE1 function would impair sodium homeostasis and as such, potentially affect the activity of other proteins associated with acid-base balance and autophagy in collecting duct cells. Therefore, any resulting effects may not be confidently attributed specifically to the mutant kAE1. With nigericin, we aimed to alter pHi while affecting the least possible other ion concentration. Due to space considerations, Figure 1 has been reorganised to include the rate of alkalinisation and pHi (panels F and G).

Reviewer #2 (Recommendations for the authors):

(1) The authors could improve the readability of this manuscript for a general audience by clarifying and summarizing the respective phenotype(s)/effect(s) of the different mutants in some kind of table in the main figures. It is hard to keep track of the different disease mutants alongside the KI mouse mutations, as the text frequently discusses multiple mutants at a time.

As requested, we added two tables (Supplementary Tables 1 & 2) in Supplementary files summarizing the data obtained in this study. We hope this will help the readership to keep track of each variant’s phenotype.

(2) The subtitle of the results section of Figure 2 should be reworded to reflect that whole kidney lysates are used for the KI mice and not the other mutants.

As requested, the title in the Results section has been modified (lines 178-179).

(3) More discussion of why the different mutants cause different strengths of phenotypes should be included.

Different variants induce different degree of functional defects as seen in Figure 1F & G. The kAE1 R295H, the only amino acid substitution in the amino-terminal cytosol causing dRTA, does not affect the transporter’s function or cells’ pHi. Therefore, this variant may cause dRTA via a different pathway than transport-defective S525F or partially inactive R589H variants that both affect pHi. Our study does not exclude that dRTA may be caused by other defects than pHi alterations, including defective proteinprotein interactions. This discussion is now included in the manuscript on lines 386-391.

Reviewer #3 (Recommendations for the authors):

In general, I found the subject matter of this manuscript interesting and of value to the scientific community. The interpretation of the data and how much it supports the conclusion that "kAE1 variants increases pHi which alters mitochondrial function and leads to reduced cellular energy levels that eventually attenuate energy-dependent autophagic pathways" is largely incomplete. There are significant concerns about the quantification of Western blot data. Additionally, including the R607H variant in the in vitro experiments would improve the interpretation and extrapolation of in vitro data to the kidney.

We apologize for the confusion with R589H and R607H variants. The R607H mutant is the murine ortholog to the human R589H dRTA variation. To clarify this, we have added this information on line 180, in addition to lines 118-119 and line 70.

Suggestions:

(1) Can an anion replacement experiment be performed in the mIMCD cells (no Cl or no HCO3) to determine that bicarbonate transport through AE1 is responsible for the reduced ATP rates in Figure 5? Inclusion of WT +dox control would be helpful to convince the reader of the effects.

Because Seahorse real-time cell metabolism ATP rates measurements require specific and patented buffers with un-specified compositions, it was not possible to modify the Cl⁻ or HCO₃⁻ content during the ATP measurement assay. All cell lines, including empty vector cells (EV) were treated with doxycycline; thus, WT + dox was already included. The empty vector cell line treated with doxycycline allowed the exclusion of specific effects of doxycycline on mitochondrial activity as a control. This is now clarified in Figure 5 legend, lines 655-656.

(2) Can the authors measure pHi in fresh kidney sections from the R607H mouse?

Unfortunately, we are not currently able to measure pHi in fresh kidney sections and although we recognize it would benefit greatly to our study, establishing a new collaboration to perform this measurement would significantly delay the publication of this work; therefore, these results will not be available for the present manuscript.

(3) Does pH 7.0 media have any effect on autophagy, as shown in Figure 3? Why was pH 6.6 selected?

The idea was to artificially acidify pHi in mutant cell lines (that have a steady state alkaline pHi) and assess whether this acidification corrects autophagy defects. We first determined that incubation in cell culture medium at pH 6.6 with 0.033 µM nigericin (final potassium concentration: 168 mM) for 2 hours provided optimal conditions, i.e. ensuring cell viability over the 2-hour period while effectively lowering intracellular pH to 6.9, as demonstrated in Supplementary Figure 1A-C.

(4) In vitro experiments should be performed on polarized cells with kAE1 properly inserted in the basolateral membrane. Experiments on subconfluent, non-polarized cells do not support the hypothesis that transport functions of AE1 initiate the cascade of events attributed to these SLC4A1 mutations.

To address this point, we have performed cell surface biotinylations on 70-80 % confluent mIMCD3 cells expressing kAE1 WT, S525F or R589H mutants and show that cell surface abundance of the mutants is not significantly different from the WT protein. This is now shown in Figure 3 A&B. As cell surface biotinylation provides a more quantitative assessment of protein cell surface abundance, we have removed the immunofluorescence images from non-polarised cells and replaced them with representative immunoblots from a cell surface biotinylation assay.

Concerns:

(1) No information about the B1 ATPase antibody used.

Now provided in Supplementary Material, ATP6V1B1 Antibody from Bicell cat#20901.

(2) No actin band in Figure 1E (as prepared).

Actin bands are provided for each blot in Figure 1D.

(3) Figures 1E and 1F are labelled wrong in the figure versus the results section.

Thank you for letting us know, this is now corrected.

(4) The cortical sections shown in Figure 4 for the KI/KI do not appear to have the morphology of a CCD. The authors may want to consider including glomeruli to convince the reader of the localization of the tubules. Same concern with Figure 5G and I. The WT image in 5G does not have the morphology of a CCD. Principal cells should be predominant, and ICs should be dispersed.

Both figures 4 and 5 have been updated with images showing glomeruli (light blue “G” on figure) with neighbour and dispersed IC staining.

(5) The quantification of LAMP1 in Figure 4 is unclear. How did the authors determine the boundary of AICs, and how did they calculate the volume of lysosomes? If a zstack was used, how are the authors sure that their 10um section includes the entire AIC?

The quantification of LAMP1 is detailed under “Image analysis”, then “Volocity” sections in Supplementary Material. The boundary of A-IC was manually detected in Volocity based on the presence of the H+-ATPase before Volocity analysis for lysosomal volume as described in the Methods.

The 10 micron sections are expected to include full AIC as well as partial AIC, but the frequency of these events should be the same between WT and variants’ sections, therefore they were all included in the analysis if cells displayed H+-ATPase signal.

(6) Figure 5: There is no description of how ATP rates are calculated from the provided traces.

We used Agilent Seahorse XF ATP rate assay kit for this experiment. In this assay, the total ATP rate is the sum of ATP production rate from both glycolysis and oxidative phosphorylation. Glycolysis releases protons in a 1:1 ratio with ATP hence the glycolytic ATP rate is calculated from the glycolytic proton efflux rate (glycoPER). GlycoPER is determined by subtracting respiration linked proton efflux from total proton efflux by inhibiting complex I and III. This information is now added to Supplementary Material, in the “Metabolic Flux analysis” section.

(7) Figure labels in Figure 5 are wrong. It seems 5H (as presented) should actually be labeled 5G. In 5H (G?), why did some cells not have any TOM20 pixel intensity for S525F and R589H variants?

Confocal image acquisition in this experiment was kept under the same settings to allow comparison between samples. Therefore, some cells show dimer fluorescence than others. From the figure 5 panels, all cells showed TOM 20 pixel intensity. Figure 5H panel has been relabelled Figure 5G.

(8) In Figure 2, the summary graphs show analysis of more samples than are visible on the included western blots. What is the rationale for this? Why does S525F have 9 samples in BafA1 while R295H only has 3 (2H)? Yet, R295H has 6 samples in 2I. In 2D, S525F has at least 9 samples. Explain.

Figure 2A-C shows representative immunoblots, among several ones independently conducted. Therefore, the final number of samples is higher than showed on Figure 2. This is now indicated in Figure 2 legend, line 603. It became clear quite early in our study that the recessive kAE1 R295H variant does not behave similarly to the other variants studied, maybe because it affects the cytosolic domain, so we did not perform as many replicates for this variant as we did for the others. However, we felt it was valuable to the research community to report the characterization of this variant and decided to keep it in our study.

(9) In general, the actin loading does not appear to be equal between samples. And some figures show the same actin blot twice (2A, C) while some show independent actin bands for LC3B and p62. Equal loading seems a fairly significant control, considering the importance of quantification in the figures.

In addition to performing protein assays, we systematically conduct immunoblot with anti-b-actin antibody to control for loading variability. When possible, two or three proteins, including actin, are detected on the same blot, when molecular weight differ enough. This sometimes results in b-actin being used as a loading control for two different proteins, as seen on Figure 2A and 2C. This is now indicated on lines 605606.

(10) In the Supplemental Figure 2, which band is being quantified for mature CTSD at 33kDa? Same for intermediate CTSD. The quantification of V-ATPase seems questionable based on the actin variance shown in the blot. Surely the ratio of the fourth sample is greater than 1.

Supplementary Figure 2 has been updated to include arrows indicating which band was selected for the quantification. After verifying the measurements of band intensities from “Image Lab” quantification software, we confirm the results, including that fourth KI/KI sample has a ratio of 0.78 (Adj Total Band Vol (Int), lanes 10). Screen shots of quantifications are attached below.

(11) Why are the experiments performed on non-confluent IMCD cells? Figure 1D shows good basolateral localization of AE1, yet the other experiments in the manuscript appear to use IMCD cells in low confluent states, without proper localization of AE1. Figure 3A shows AE1 dispersed throughout the cytoplasm. Why have the authors decided to study the effects of an anion exchanger without it being properly localized to the basolateral membrane? Shouldn't all experiments be performed in polarized IMCDs? If AE1 isnt properly in the membrane, and the cells do not have defined apico-basolateral polarity, then what role can AE1-mediated intracellular pH change have on the results of the experiments? Were the pHi experiments in 3E performed on polarized cells? Or even 1F?

To address this point, we have performed cell surface biotinylations on 70-80 % confluent mIMCD3 cells expressing kAE1 WT, S525F or R589H mutants and show that cell surface abundance of the mutants is not significantly different from the WT protein. This is now shown in Figure 3A & B. As it provides a more quantitative assessment of protein cell surface abundance, we have removed the immunofluorescence images from non-polarised cells and replaced them with a representative immunoblot from a cell surface biotinylation assay.

(12) As mentioned in the public comments, how is the ratio A/(A+B) greater than 1? With A and B > 0. In Figure 3, the data is reasonable, but in Figure 2, the data is simply impossible. What is the explanation for this phenomenon? Why was this presentation of data approved? Is it supposedly a fold of WT, like 2K and 2L? Is the reader also to believe that total LC3B is 2-fold greater in KI/KI mice, as shown in 2K? My eyes, though not densitometry equipment, cannot confirm this. The actin bands are not equal. Yet again, there are 4 lanes of KI/KI mice, but the quantification shows 5 samples.

The ratios in figure 2D, 2F, 2H and 2L have been re-calculated and corrected. As indicated above, immunoblots are representative and quantification of additional blots has been included in the graphs.

(12) Spelling error Figure 4B: cels.

Corrected

References

(1) Mumtaz, R. et al. Intercalated Cell Depletion and Vacuolar H+-ATPase Mistargeting in an Ae1 R607H Knockin Model. Journal of the American Society of Nephrology 28, 1507–1520 (2017).

Version 2

Abstract

Distal renal tubular acidosis (dRTA) is a disorder characterized by the inability of the collecting duct system to secrete acids during metabolic acidosis. The pathophysiology of dominant or recessive SLC4A1 variant related dRTA has been linked with the mis trafficking defect of mutant kAE1 protein. However, in vivo studies in kAE1 R607H dRTA mice and humans have revealed a complex pathophysiology implicating a loss of kAE1-expressing intercalated cells and intracellular relocation of the H+-ATPase in the remaining type-A intercalated cells. These cells also displayed accumulation of ubiquitin and p62 autophagy markers. The highly active transport properties of collecting duct cells require the maintenance of cellular energy and homeostasis, a process dependent on intracellular pH. Therefore, we hypothesized that the expression of dRTA variants affect intracellular pH and autophagy pathways. In this study, we report the characterization of newly identified dRTA variants and provide evidence of abnormal autophagy and degradative pathways in mouse inner medullary collecting duct cells and kidneys from mice expressing kAE1 R607H dRTA mutant protein. We show that reduced transport activity of the kAE1 variants correlated with increased cytosolic pH, reduced ATP synthesis, attenuated downstream autophagic pathways pertaining to the fusion of autophagosomes and lysosomes and/or lysosomal degradative activity. Our study elucidated a close relationship between the expression of defective kAE1 proteins, reduced mitochondrial activity and decreased autophagy and protein degradative flux.

Introduction

Distal renal tubular acidosis (dRTA) is a disorder characterized by the inability of the collecting duct system to secrete acids during metabolic acidosis1. In addition to hyperchloremic metabolic acidosis, patients with this disease can present with hypokalemia, kidney stones, urinary sodium waste and difficulty to thrive. Expression of pathogenic variants in the ATP6V0A4, ATP6V1B1, FOXI1, WDR72 and SLC4A1 genes are the usual genetic aetiologies2–4. The SLC4A1 gene encodes the anion exchanger 1 (AE1) protein, which is an electroneutral chloride/bicarbonate exchanger5. It exists in two forms: a 911 amino acid erythroid isoform known to interact with erythroid cytoskeletal proteins and participate in red cell respiration and integrity, and a 65 amino acid (NH2-terminal) truncated isoform primarily found in the basolateral membrane of renal type A intercalated cells (A-IC)6,7 and podocytes8. This isoform participates in bicarbonate reabsorption and through its physical and functional interaction with the cytosolic carbonic anhydrase II and apical H+-ATPase, it supports apical proton export and urine acidification9.

Pathogenic variants in the SLC4A1 gene can result in either red cell defects (such as Southeast Asian ovalocytosis10,11 and hereditary spherocytosis12), renal cell defects (dRTA10) or both in patients with homozygous (Band 3 Coimbra and Band 3 Courcouronnes13,14) or compound heterozygous variants15,16. Renal SLC4A1 disease-causing variants have only been found in the transmembrane domain - where it could impact protein structure and its transport function- or in the short carboxyl domain – where it possibly affects protein-protein interactions. The pathophysiology of dominant or recessive SLC4A1 related dRTA (hereafter named dRTA) has originally been linked with the mis-trafficking defect of mutant kAE1 protein10,17. However, recent in vivo studies in R607H (orthologous to human R589H dRTA variant) and L919X knockin mice and dRTA patients have revealed a complex pathophysiology where kAE1-expressing intercalated cells were lost, and in the remaining type-A intercalated cells, the H+-ATPase relocated intracellularly and accumulated autophagy marker p62 and ubiquitin-positive material18.

The highly active transport properties of collecting duct cells require the tight maintenance of cellular energy and homeostasis. The autophagy-mediated turnover of damaged organelles is necessary for protecting collecting duct cells as in most renal cells19. The chloride/bicarbonate exchange function of kAE1 in A-ICs confers a pivotal role in pH homeostasis and thus is a major contributor to cellular homeostasis. kAE1 protein interacts with several proteins such as integrin-like kinase (ILK), adaptor-related protein complex 1, 3 & 4 (AP-1, AP-3 & AP-4 mu1A), transmembrane protein 139 (TMEM139), kinesin family member 3B (KIF3B), and clathrin among others7,20–22 that support protein stability and trafficking. It also interacts with homeostatic proteins including the glycolytic enzyme glyceraldehyde-3-phosphate dehydrogenase (GAPDH)23 and the antioxidant enzyme peroxiredoxin 6 (PRDX 6)24, which play major roles in cellular energy metabolism and oxidative stress response, respectively.

In this study, we report the characterization of newly identified dRTA genetic variations and provide evidence of abnormal autophagy and degradative pathways in cells and kidneys from mice expressing dRTA mutant kAE1 proteins.

Concise Methods

Ethics Approval

This study was conducted in accordance with all national and institutional animal care guidelines and approved by the University of Alberta’s Animal Care and Use Committee (AUP #1277).

Newly identified SLC4A1 variations from dRTA patients

The patient carrying the R295H mutation was a boy carrying the variation in the homozygous state, whose genetic diagnosis was made at the age of 5, following growth retardation of −2 SD for both weight and height, with bicarbonate at 18 mmol/L, potassium at 2.8 mmol/L, chloridemia at 94 mmol/L, calcemia at 2.55 mmol/L, and a urinary pH of 7.5. He also had a history of a pyeloureteral junction syndrome that was surgically managed at the age of 3. The R295H dRTA variation is a nonsense homozygous substitution characterized by a replacement of guanine (G) on position 884 by adenine (A) in the coding sequence. It has an allelic frequency of 0.14 % in the European population.

For the Y413H variant, the patient was a female diagnosed at 1 month old, with an urinary pH of 7.5, evidence of nephrocalcinosis and failure to thrive. The S525F variant has been previously reported25, but in brief, the patient was a 13 year old female with plasma pH of 7.25, plasma bicarbonate at 15.3 mmol/L who also presented with polyurethral junction syndrome, nephrocalcinosis and nephrolithiasis since childhood. The Y413H and S525F dRTA variations are nonsense heterozygous substitutions characterized by a replacement of thymine (T) at position 1237 by cytosine (C) and a replacement of cytosine (C) at position 1574 by thymine (T), respectively. The R589H dRTA variation has been previously described18. No follow up data were available for all patients.

Mice

Transgenic mice carrying the R607H knockin (KI) mutation (murine equivalent to human R589H mutation) were previously described18. Homozygous mice used throughout the study display incomplete dRTA with alkaline urine without metabolic acidosis at baseline as previously reported18. Homozygous mice or wild-type (WT) littermates were fed a standard rodent chow (Picolab® Rodent Diet 20 # 5053, LabDiet, ST. Louis, MO, USA) or for Figure 5 (H-K), a salt-depleted diet with acid challenge as previously reported26 with adequate and constant water supply, and maintained on a 12-hour light and dark cycle throughout their lifespan.

Statistical analysis

All the experiments were independently repeated a minimum of three times. Experimental results were analyzed using the GraphPad Prism software and are summarized as mean ± SEM. All statistical comparisons were made using unpaired Student’s t test or one/two-way ANOVA followed by a post-hoc test as indicated in figure legends. A P-value less than 0.05 was considered statistically significant. All datasets were assessed for normality, and outliers identified by Prism were excluded.

Detailed Material and Methods are listed in the Supplementary Material.

Results

The dRTA kAE1 variants traffic to the basolateral membrane but have reduced transport activity in mIMCD3 cells

We first characterized three newly identified dRTA mutations and compared them with kAE1 WT or previously characterized kAE1 R589H mutant. Figure 1A depicts the alpha fold predicted structure of kAE1 showing amino acids mutated in each kAE1 mutant. kAE1 R295H is a recessively inherited substitution in the N-terminal cytosolic domain of the protein. kAE1 Y413H is a dominantly inherited substitution in transmembrane domain (TM) 1 of the core domain. In the gate domain, dominantly inherited S525F and R589H substitutions occur in TM 5 and TM 6, respectively. We generated the newly identified dRTA variant cDNAs and expressed them or kAE1 WT in mIMCD3 cells. As seen on Figure 1B, kAE1 R295H, S525F and R589H variants display 2 typical bands similar to kAE1 WT. The top band (open circle) encompasses kAE1 proteins carrying a complex oligosaccharide that have reached the Golgi and beyond, while the bottom band (black circle) corresponds to high mannose-carrying kAE1 proteins located in the endoplasmic reticulum. However, kAE1 Y413H mutant bands intensity was overall weaker than WT and displayed a predominant single band aligned with high mannose-carrying kAE1 proteins. These results indicate that the 3 newly described dRTA mutants are successfully expressed in mIMCD3 cells. We next localized these mutants by immunofluorescence in polarized mIMCD3 cells. Both kAE1 WT and mutants appropriately co-localized with basolateral membrane marker beta-catenin in polarized mIMCD3 cells (Figure 1C). kAE1 R589H location has previously been reported at the basolateral membrane in polarized mIMCD3 cells18. However, staining for kAE1 Y413H was again weaker and seemed more intracellular than other mutants. To address a possible premature degradation of this variant, we measured its lifetime and observed that its degradation begins 6 hours post-synthesis while kAE1 WT abundance remained stable for 24 hours (Figure 1D). Finally, to assess the transport activity of the mutants, we examined the steady-state cytosolic pH (pHi) and rate of intracellular alkalinisation of mIMCD3 cells expressing kAE1 WT or mutants (Figure 1E-G). Using BCECF-AM, we observed that the steady state intracellular pH (pHi) of kAE1 mutant cells was more alkaline than WT cells (except for kAE1 R295H cell pHi, which has a similar trend but is not significantly different from WT) (Figure 1F), in agreement with rate of intracellular alkalinisation (Figure 1G). Note that transport activity, measured as rate of alkalinisation, is measured by reverting kAE1 exchange activity from bicarbonate export to import (see Methods, protocol for transport assay), hence a reduced alkalinisation rate is observed in cells expressing defective kAE1 protein compared to WT (panel G). Overall, these results indicate that except for kAE1 Y413H mutant, the other newly described variants are expressed and traffic to the basolateral membrane of polarized mIMCD3 cells, similar to the previously published kAE1 R589H mutant. Given that the premature degradation of kAE1 Y413H mutant likely explains the dRTA phenotype, we did not perform further assays on cells expressing this protein.

kAE1 R295H, Y413H, S525F and R589H dRTA mutants are either dysfunctional or prematurely degraded.

(A) Alpha Fold predicted structure of the kidney isoform of Band 3 anion exchanger 1 (kAE1) with core and gate domains highlighted. The dRTA kAE1 mutation sites are coloured in blue with line extensions detailing specific amino acids mutated. (B) Immunoblot showing expression of kAE1 WT, R295H, Y413H, S525F and R589H and corresponding actin band in mIMCD3 cells treated with and without doxycycline for 24 hrs. Mouse anti-HA antibody was used to detect kAE1-HA, top (open circle) and bottom bands (closed circle) correspond to kAE1 carrying complex and high mannose oligosaccharides, respectively. (C) Immunostaining of kAE1 WT or mutant (red) and β-catenin (green) in polarized mIMCD3 cells. Scale bar = 10 μm. Red = kAE1, Green = ß-Catenin. (D) Immunoblot of cycloheximide (CHX) chase assay with corresponding actin in kAE1 mIMCD3 WT and Y413H cells showing the degradation of kAE1 Y413H after 3hrs CHX incubation. (E) Cartoon depicting the transporter activity and expected changes in pHi in cells expressing kAE1 WT (left) or inactive mutant (right). (F) Graphical representation of intracellular pH (pHi) measurement of mIMCD3 kAE1 WT, R295H, Y413H, S525F and R589H cells. Error bars correspond to mean ± SEM, n = minimum 32. *P<0.05, **P<0.01 using one-way ANOVA followed by a Dunnett’s post-hoc test. (G) Rate of intracellular alkalinisation in WT or mutant mIMCD3 cells normalized to WT+ Dox. **** indicates P < 0.0001 using one-way ANOVA followed by a Dunnett’s post-hoc test. Error bars correspond to mean ± SEM, n= minimum 4

Autophagy processes are altered in mIMC3 cells expressing the kAE1 R295H, S525F and R589H dRTA mutants and in R607H knockin kidney lysates

In mice expressing kAE1 R607H (the murine equivalent to human dRTA R589H substitution), a striking reduction in type-A intercalated cells was noted and in the remaining cells, autophagy marker p62 and ubiquitin accumulated in these abnormally enlarged cells18. We therefore investigated the autophagy machinery in dRTA mutant mIMCD3 cells and in R607H knockin (KI) mice. We first examined the ratio of autophagosome marker LC3BII protein relative to the total intensities of LC3BI and LC3BII as well as p62 levels. These experiments were performed at steady state, upon autophagy induction by starvation, or after autophagy inhibition by Bafilomycin (Baf) A1 (Figure 2A-C). Figure 2A – I shows a consistent increase in the ratio of LC3B II to total LC3B (I + II) in the mutant cells at steady state (panel D), with starvation (panel F) and with Baf A1 (panel H) except for the kAE1 R295H mutant which was not significantly different from WT. This suggests an altered autophagy process in the mutant cells, in agreement with preliminary findings from Mumtaz and colleagues18. There was no significant difference in p62 abundance in mutant cells compared to WT at steady state and with starvation (Figure 2E&G). However, with Baf A1, R589H mutant cells had significantly lower p62 abundance compared to WT (Figure 2I). To confirm these findings, we next assessed the abundance of these markers in whole kidney lysates from the kAE1 R607H KI mice. We observed that the total LC3B abundance was significantly higher in homozygous KI mice (KI/KI) compared to WT, with no difference in p62 (Figure 2J-L). Overall, these results suggest an abnormal autophagy in mutant mIMCD3 cells and in kidneys of R607H KI mice. Given the lack of difference in phenotype between the recessive kAE1 R295H and kAE1 WT mIMCD3 cells, we focused the subsequent experiments on dominant kAE1 S525F and R589H mutant cells. Further investigations will be needed to understand the pathophysiology associated with the kAE1 R295H novel variant.

Autophagy is upregulated in dRTA kAE1 mutants in vitro and in vivo.

(A-C) Representative immunoblots of LC3B and p62 with corresponding actin abundance in kAE1 WT, R295H, S525F and R589H mIMCD3 cells at steady state, under starvation (Starv) or 400 nM Bafilomycin A1 (Baf) treatment. Note that p62 and LC3B were detected on the same blot for panel A and C, therefore the same actin blot is shown for both panels. (D-I) Quantification of all immunoblots showing the ratio of LC3B II to total LC3B and p62. Error bars correspond to mean ± SEM, n= 3-8. * P<0.05, ** P<0.01, *** P<0.005, ****P < 0.001 using one-way ANOVA followed by a Tukey’s post-hoc test. Immunoblots (J) and quantification (K, L) of LC3B and p62 abundance in kAE1 R607H KI mouse whole kidney lysates. Error bars correspond to mean ± SEM, n= minimum 5. ***P < 0.005 using one-way ANOVA followed by a Tukey’s post-hoc test.

Late autophagy steps are blocked in dRTA kAE1 mutant-expressing cells due to their alkaline intracellular pH

Given these preliminary findings of abnormal autophagy in dRTA mutant cells, we examined in more detail their autophagy machinery by transiently transfecting them with the eGFP-RFP-LC3 construct and monitoring the rate of autophagosome and autolysosome formation. In this assay, the green fluorescent protein (eGFP) fused to LC3 is quenched in the acidic environment of the autolysosome, while both eGFP and red fluorescent protein (RFP) fluoresce in vesicles in the neutral lumen of the autophagosome27,28. To avoid the poor efficiency of transient transfection in polarized cells, experiments were conducted in sub-confluent cells pending that the mutant proteins are present at the plasma membrane. We therefore performed cell surface biotinylations on 70-80 % sub-confluent cells, which confirmed a robust plasma membrane abundance of both kAE1 R589H and S525F that was not significantly different from the WT protein (Figure 3A&B). Therefore, we next assessed the efficiency of the autophagy machinery in WT or dRTA mutant-expressing cells (Figure 3C-F). Focusing on cells expressing either kAE1 WT, kAE1 S525F or R589H, we quantified the number RFP+ (red, acidic autolysosomes) and double eGFP+/RFP+ (yellow, not acidic autophagosomes) vesicles. kAE1 S525F mutant cells have significantly more autophagosomes than WT counterparts (Figure 3D), and both kAE1 S525F and R589H have significantly more autolysosomes than WT (Figure 3E). Figure 3F shows that both mutants had significantly more autophagosomes (yellow) than autolysosomes (red). This finding suggests an upregulation of autophagy and inhibition in the downstream steps of the process that involves the fusion of autophagosomes with the lysosome29. As autolysosomes require luminal v-H+-ATPase-dependent acidification to efficiently clear cell debris 30, the higher intracellular pH seen in mutant cells (Figure 1F) may be detrimental to v-H+-ATPase full activity and impair proper autolysosomal acidification31. We therefore wondered whether chemically acidifying the pHi in mutant cells would rescue the autophagy machinery (Figure 3G-K). We first determined that incubation of mIMCD3 cells in 0.033 μM nigericin in cell culture medium at pH 6.6 for 2 hours acidified cytosolic pH to 6.9 without causing cell death (Supplementary Figure 1A-C). Next, we observed that chemically reducing pHi to 6.9 in mutant expressing cells reduced the ratio of LC3B II to total LC3B (Figure 3J) and the abundance of lysosomal-associated membrane protein 1 (LAMP1) in R589H cells (Figure 3 K) to levels similar to WT cells at steady state. These findings suggest that abnormal autophagy in the mutant cells may be caused by their alkaline pHi, resulting from a reduced anion exchange activity of the mutant kAE1 protein.

dRTA kAE1 mutants have more alkaline steady-state intracellular pH and altered autophagy flux.

(A) Representative immunoblots of cell surface biotinylation experiments in mIMCD3 cells expressing kAE1 WT or mutants (top panels), with control staining of plasma membrane marker Na+/K+-ATPase (middle panel) and intracellular marker actin (bottom panel). (B) Quantification of 3 independent cell surface biotinylation experiments. Data are represented by a single representative blot for each variant. N.s., not significant using one-way ANOVA. Error bars correspond to mean ± SEM, n=3. (C) Immunofluorescence staining in eGFP-RFP-LC3 transfected mIMCD3 cells expressing kAE1. GFP = green, RFP = red, kAE1= cyan, nuclei = dark blue (merge only). Scale bar = 2 μm. Graphical representation of number of yellow (autophagosomes) (D) and red (autolysosomes) (E) puncta per cell expressing kAE1. Error bars correspond to mean ± SEM, n= minimum 32. ** P<0.01, *** P<0.005, ****P < 0.001 using one-way ANOVA followed by a Tukey’s post-hoc test. (F) Grouped graph of the number of yellow (autophagosomes) and red (autolysosomes) puncta per cell expressing kAE1, respectively. Note that the statistical analysis displayed only compared yellow and red groups for simplification. Error bars correspond to mean ± SEM, n= minimum 32. ** P<0.01, ****P < 0.001 using two-way ANOVA followed by a Sidak’s post-hoc test. (G-I) Immunoblot of LC3B, LAMP1 and actin in kAE1 WT, S525F and R589H mIMCD3 cells at steady state and under chemically reduced intracellular pH conditions. Graphical representation of the ratio of LC3B II to total LC3B ratio (J) or LAMP 1 (K) at steady state versus at low pHi in mIMCD3 kAE1 WT, S525F and R589H. Black circles indicate steady state cells and triangles indicate low pHi cells. Error bars correspond to mean ± SEM, n= 3. ** indicates P<0.01 using two-way ANOVA followed by a Sidak’s post-hoc test.

mIMCD3 cells expressing dRTA kAE1 mutants and R607H KI kidney tissues have abnormal lysosome number and size

The accumulation of autophagosomes and autolysosomes as seen above may suggest one or a combination of the following: an inability of autophagosomes to fuse with lysosomes and/ or a defect in lysosomal degradative activity in the mutant cells19,32. We first examined the lysosomal degradative activity by assessing lysosomal protease Cathepsin B activity using Magic Red staining, a probe that fluoresces upon lysosomal protease cleavage (Figure 4A-B). In agreement with increased RFP+ vesicles (Figure 3E), the kAE1 S525F mutant cells had a significantly higher number of Magic Red positive vesicles than WT whereas the kAE1 R589H mutant cells had significantly larger Magic Red positive vesicles, suggesting an accumulation of undigested material33. To validate this finding in vivo, we performed immunostaining and quantified LAMP1 positive staining in ß1 ATPase positive cells (a marker of ICs) in WT and R607H KI mouse kidney sections. The KI mice showed significantly more and larger LAMP1 positive vesicles compared to WT mice in both cortex and medulla (Figure 4 C-F). We probed further into the lysosomal activity by quantifying lysosomal protease Cathepsin D (immature, intermediate and mature) protein abundance by immunoblot in isolated primary murine A-ICs. Although the abundance of immature and intermediate cathepsin D did not differ between genotypes, the KI mice showed a significantly decreased abundance of mature cathepsin D (Supplementary Figure 2). Thus, in line with in vitro findings, A-IC from homozygous R607H KI mice display relatively more and larger lysosomes with reduced active protease abundance than WT littermates, suggesting a lysosomal defect in the dRTA kAE1 mutant cells.

dRTA kAE1 mutants have bigger or more lysosomes than WT in vitro and in vivo.

(A) Immunofluorescence images of kAE1 WT, S525F and R589H mIMCD3 cells incubated with Magic Red substrate for 1hr at 37°C in the dark. Green = kAE1, magenta = active lysosomes, blue = nuclei. Scale bar =16 µm. (B) Graphical representation of number and size of active lysosomes per cell. Error bars correspond to mean ± SEM, n= minimum 30. **** P<0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. Immunofluorescence images of LAMP 1 and ß1 ATPase in kidney cortex (C) or medulla (E) from kAE1 WT and R607H KI mice. Blue = nuclei, magenta = LAMP 1 (lysosomes), Yellow = ß1 ATPase, light blue “G” indicates the location of a glomerulus. Scale bar = 8μm. Graphical representation of the number and volume of LAMP1 vesicles in ß1 ATPase positive cells in the kidney cortex (D) or medulla (F) of WT or R607H KI mice. Error bars correspond to mean ± SEM, n= 60. **** P<0.001 using Student’s t-test.

dRTA kAE1 mutant cells have lower ATP production rate and abnormal mitochondrial content.

Oxygen consumption rate (OCR) (A) and Extra Cellular Acidification Rate (ECAR) (B) of empty vector (EV) transfected cells, kAE1 WT, S525F or R589H mIMCD3 cells analyzed in a Seahorse XFe96 Extracellular Flux Analyzer with the ATP Rate Assay Test Kit. All cell lines, including EV transfected cells, were incubated with doxycycline, to eliminate a potential effect of doxycycline on measurements. (C) Graphical representation of the combination of ATP production rates from mitochondrial respiration (mitoATP) and glycolysis (glycoATP) of kAE1 WT, S525F and R589H mIMCD3 cells measured in real-time following sequential injections of oligomycin and Rotenone +Antimycin A. Error bars correspond to mean ± SEM, n= minimum. * p < 0.05, ****p < 0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. Graphical representations of mitochondrial respiration (D) and glycolytic ATP production (E) in kAE1 WT, S525F and R589H mIMCD3 cells. Error bars correspond to mean ± SEM, n= minimum 8. **P<0.01, ****p < 0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. (F) Immunofluorescence staining of TOM 20 and kAE1 in kAE1 WT, S525F and R589H mIMCD3 cells. Blue = nuclei, Magenta = TOM 20, Green = kAE1. Scale bar = 8 µm. (G) Graphical representation of total TOM 20 fluorescence intensity per cell expressing kAE1. Error bars correspond to mean ± SEM, n= minimum 40. ***p < 0.001, ****p < 0.0001 using one-way ANOVA followed by Tukey’s post-hoc test. Immunofluorescence images of TOM 20 and ß1 ATPase in kidney cortex (H) or medulla (J) of kAE1 R607H WT and KI mice exposed to a salt depleted diet with an acid challenge26. Blue = nuclei, magenta = TOM 20 (mitochondria), Yellow = ß1 ATPase, light blue “G” indicates the location of a glomerulus. Scale bar = 8μm. Graphical representation of the total TOM 20 fluorescence intensity in ß1 ATPase positive cells in the cortex (I) or medulla (K) of the kidney. Error bars correspond to mean ± SEM, n= 90. * p < 0.05, ****p < 0.0001 using Student’s t-test.

dRTA kAE1 mutant cells have lower ATP production rate and abnormal mitochondrial content

Lysosomal degradation is highly dependent on a low luminal pH generated in part by the vacuolar-type H+-ATPase34 whose activity depends on ATP hydrolysis. We therefore analysed the ATP production rate in mIMCD3 cells, specifically glycolysis and oxidative phosphorylation. We measured the oxygen consumption rate (OCR) (Figure 5A) and extracellular acidification rate (ECAR) (Figure 5B) in empty vector transfected, kAE1 WT or mutant cells. Both kAE1 S525F and R589H mutant cells had a lower ATP production rate compared to WT (Figure 5C). More specifically, the R589H mutant cells had a lower mitochondrial ATP production rate (Figure 5D) whereas the kAE1 S525F mutant cells exhibited a lower glycolytic ATP production rate (Figure 5E). With the mitochondria being the major ATP-producing organelles35, we assessed mitochondrial content by immunostaining of translocase of the outer membrane 20 (TOM20) both in vitro and in vivo. Both kAE1 S525F and R589H mutant cells have higher mitochondrial content compared to WT as determined by total overall intensity of TOM20 positive puncta (Figure 5F&G). In line with this result, although a decreased fluorescence intensity was observed in the cortex, there was a significantly higher TOM20 fluorescence intensity in medullary kidneys of homozygous R607H KI mutant mice compared to WT littermates (Figure 5H-K).

Discussion

In this study, we characterized three newly identified dRTA-causing kAE1 variations. Combining in vivo and in vitro studies, we demonstrated that reduced transport activity of the kAE1 mutants correlated with increased cytosolic pH, reduced ATP synthesis, attenuated downstream autophagic pathways and lysosomal dysfunction, pertaining to the fusion of autophagosomes and lysosomes and/ or lysosomal degradative activity (See summary in Supplementary Tables 1 and 2).

In line with previous observations in mIMCD3 cells18, the kAE1 R295H, Y413H and S525F mutants were properly localized to the basolateral membrane after polarization. When expressed in Madin-Darby canine kidney (MDCK) cells, other dRTA-causing kAE1 mutants such as dRTA R602H, G701D, V488M, deltaV850 variants exhibited a plasma membrane trafficking defect10,36,37. In contrast, the kAE1 R589H mutant was correctly targeted to the plasma membrane18. Functionally, cells expressing the Y413H and S525F mutants exhibit about 60 % reduction in chloride/bicarbonate exchange activity compared to kAE1 WT, similar to the previously characterized recessive G701D mutant but unlike the R295H mutant38. Therefore, the mechanism causing dRTA remains unclear in the case of the newly identified recessive R295H variant. Additionally, the kAE1 Y413H variant exhibited a shorter half-life than WT counterpart, likely explaining dRTA, and thus was not further investigated. These findings add to the growing list of SLC4A1 gene variations causing dRTA.

We next investigated the roots of the altered autophagy briefly reported in R607H KI mice18 using mIMCD3 cells and whole kidney lysates. In kidneys of the KI mice, a decrease in the number of A-IC, accumulation of p62 and ubiquitinylated proteins and enlarged remaining A-ICs, suggested altered autophagy processes in dRTA mutant cells and in homozygous R607H KI mice18,38. During the autophagy process, LC3B I (a marker for autophagosomes) is converted to lipidated LC3B II39 and p62 aggregates to facilitate the degradation of ubiquitinated proteins within the autophagosome complex40. In mIMCD3 cells expressing dRTA kAE1 S525F and R589H variants, LC3B lipidation was increased compared to WT, an increase that persisted with both Baf A1 treatment and starvation. Although opposite effects were expected under inhibition or induction of autophagy, such similar effect has been previously described. In the proximal tubule of obese mice, LC3B accumulation indicating a stagnated autophagy flux, was observed with both chloroquine treatment and 24 hour starvation41. Similarly, in NRK-52E cells, a disruption of the autophagy machinery in high cadmium-stressed cells resulted in LC3B II accumulation under either Baf A1 or rapamycin (an autophagy inducer) treatment42. Although LC3B II is elevated during both increased autophagy flux and disrupted autophagy, we did not observe significant accumulation of p62 in mIMCD3 cells expressing dRTA mutants. p62 is specifically a marker of autophagy-mediated protein clearance43–45. Therefore, p62 accumulation in R607H KI mouse kidney sections strongly suggests a compromised autophagy-mediated clearance while the increased LC3B lipidation without significant changes in p62 in mIMCD3 cells points towards either an increased autophagic flux and/or a disrupted autophagy.

To obtain a clearer picture of the precise autophagic pathway altered in the dRTA mutants, we probed further into the different stages of autophagy and autophagy flux. We noted an accumulation of autophagosomes and autolysosomes in the S525F and R589H mutant cells. This was recapitulated in the R607H KI mice which showed significantly more and larger LAMP 1 positive vesicles in both kidney cortex and medulla, suggesting a blockage in late steps of autophagy flux in both dRTA mutant cells and KI mice. Such blockage has been implicated in the pathophysiology of several diseases. In lysosomal storage disease, lysosome accumulation in proximal tubule cells is a key component in the pathways mediating epithelial dysfunction19. In this study, restoring autophagy flux attenuated disease progression. Another study in SK-N-SH, RT4-D6P2T and HeLa cells implicated autophagosome and lysosome accumulation in cellular toxicity associated with neurodegenerative diseases32. In agreement with altered lysosomal function, we also noted a greater abundance and size of active cathepsin B lysosomal protease vesicles in mIMCD3 cells. Increased cathepsin B activity affects lysosomal biogenesis, autophagy initiation and cellular homeostasis46. In the renal context, cathepsin B knockout mice demonstrated a higher resistance and quicker recovery from glomerular damage47, suggesting that cathepsin B accumulation may be detrimental to the cells. Increased cathepsin B abundance in the dRTA mutant cells also correlates with the accumulation of lysosomes. Overall, these findings suggest that the pathogenesis of dRTA in our in vitro and in vivo models involves an inhibition of autophagy flux at the downstream steps involving autophagosome-lysosome fusion and lysosomal protein clearance 29,33,48,49.

The question remained as to how these dRTA variants altered autophagy. The SLC4A1-3 gene family that includes AE1 are regulators of intracellular pH in different cell types50–52. The reduced anion exchange activity in mIMCD3 cells expressing the R589H and S525F variants expectedly correlated with an increased pHi compared to WT counterparts. Given that pHi variations can impact autophagy31,34,53–55, we wondered whether this was the case for cells expressing kAE1 dRTA mutants. We observed that an acidic pHi in mutant cells restored autophagy levels similar to WT. Previous studies reported more perinuclear localization of lysosomes and autophagosome-lysosome fusion in cells with an increased pHi53,54. Although not examined in our study, an increase in intracellular pH due to starvation decreased the levels of lysosomal kinesin superfamily member KIF2 and ADP-ribosylation factor-like 8B (ARL8), which are responsible for redistributing lysosomes to the cell periphery. This reduction subsequently inhibited mTORC1, resulting in increased autophagosome synthesis and autophagosome-lysosome fusion53.

While an alkaline cytosolic pH partially explains the impairment in autophagy flux, a close relationship also exists between intracellular pH and cellular energy dynamics and metabolic stress, all of which are key regulators of autophagy56. Therefore, it was plausible that kAE1 variant-induced autophagy dysregulation occurs through a signalling pathway akin to that of energy deprivation-induced autophagy. This hypothesis is supported by our findings of reduced ATP production rate in dRTA kAE1 S525F and R589H mutant cells compared to kAE1 WT cells. We also found that both kAE1 S525F and R589H mutant cells have higher mitochondrial content compared to kAE1 WT cells. The increased mitochondrial content coupled with low ATP point towards improper mitochondrial function in dRTA variant cells. Low ATP levels as seen in dRTA mutant cells may impair autophagy as was shown in human RPE cells57. ATP reduction in RPE cells led to complex mitochondrial changes such as structural disorganization, enzyme activity decline, and oxidative damage to mitochondrial components and DNA57. Similarly, in pancreatic islet cells, an alkaline pHi led to increased uptake of phosphate by mitochondria, accelerating the production of superoxide, promoting mitochondrial permeability transition, and inducing translational attenuation due to endoplasmic reticulum stress, ultimately impairing insulin secretion58. Overall, our data support that expression of the kAE1 variants increases pHi which alters mitochondrial function and leads to reduced cellular energy levels that eventually attenuates energy-dependent autophagic pathways including autophagosome-lysosome fusion and lysosomal protein clearance.

In light of these observations, we postulated that correcting the alkaline pHi of dRTA mutant-expressing mIMCD3 cells will alleviate this blockage in autophagy flux. We observed that a chemically engineered pHi of 6.959 reduced LC3B II accumulation and LAMP1 abundance in mIMCD3 mutant cells to expression levels similar to that of WT cells at baseline. This suggests that a chemically reduced pHi facilitated protein clearance in the two dRTA mutant cells31,53 as noted in other studies. In one such study, treatment of SH-SY5Y cells with FCCP and nigericin also acidified intracellular pH and triggered autophagy and mitophagy31. Similarly, acid loading in proximal tubular cells under chronic metabolic acidosis showed increased autophagic flux and mitophagy60. These findings are in line with our results and establish a link between altered autophagy flux and the alkaline pHi of dRTA variant cells. Overall, our study provides one pathway (altered pHi) by which dRTA may arise. However, different variants induce different degree of functional defects as seen in Figure 1F & G. The kAE1 R295H, the only reported amino acid substitution in the amino-terminal cytosol causing dRTA, does not affect the transporter’s function or pHi. Therefore, this variant may cause dRTA via a different pathway, for example defective protein-protein interactions, than transport-defective S525F or partially inactive R589H variants.

In conclusion, our study established a strong relationship between the expression of defective kAE1 proteins, reduced mitochondrial activity, decreased autophagy and impaired protein degradative flux. Whether this abnormal degradative pathway explains the premature loss of A-IC will need to be elucidated in further studies.